A GPU Sonar Simulator for Automatic Target Recognition

### 2014-05-23 Simulator/article.tex
More on simulator article

Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}. Most implementations have however prioritized accurate acoustic modeling at the expense of execution speed. One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b} which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering. However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.

@@

@@

Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$. There are two main types of projection, perspective and orthographic. Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image. Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1]. This will render all objects equally large whether they are in the distance or up close. Its projection matrix is defined as
\begin{align}
\P = \left(\begin{matrix}
\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
0              &  0              &  0              &  1
\end{matrix}\right)
\end{align}
where $l,r,b,t,n$ and $f$ are the left, right, bottom, top, near and far boundaries of the projection plane, respectively. They define the window through which we observe our scene. We want this adapted to perfectly fit the specified image size. This is straightforward for left and right since it matches the $x$-axis
\begin{align}
l &= x_\text{min}  &  r &= x_\text{max}
\end{align}
For the near and far limits one simply needs to make sure the scene will fit in it
\begin{align}
b &=   &  t &= \\
f &= 0.1   &  n &= 200 
\end{align}
How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
\begin{align}
\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
\end{align}
and the distance from the image origin to its $y$-boundaries is given as
\begin{align}
\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
\end{align}
To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
\begin{align}
\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
\end{align}

@@

The ``camera image'' rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be. It does, however, tell us what parts of the scene that are visible from the sonar. OpenGL can also produce a depth map that reveals the distance to each of image pixels. This information can be converted to a ranged sonar-like image by simply adding up all the intensity values that share the same depth for each range line. We perform this computation in OpenCL, which allow general purpose programming on GPUs and can interoperate with OpenGL quite nicely. This way we keep all the calculations on the GPU.

@@

>>>>

Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}. However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed. One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b} which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering. However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.

\begin{itemize}
\item OpenGL does not support scattered writes. OpenCL does.
\end{itemize}

@@

\begin{itemize}
\item Load arbitrary complex models
\item Simulate sonar with OpenGL
\item Speed considerations
\end{itemize}

@@

Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$. There are two main types of projection, perspective and orthographic. Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image. Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1]. This will render all objects equally large whether they are in the distance or up close. Its projection matrix is defined as
\begin{align}
\P = \left(\begin{matrix}
\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
0              &  0              &  0              &  1
\end{matrix}\right)
\end{align}
where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene. We need these parameters set They define the window through which we observe our scene. We want this adapted to perfectly fit the specified image size. This is straightforward for left and right since it matches the $x$-axis
\begin{align}
l &= x_\text{min}  &  r &= x_\text{max}
\end{align}
For the near and far limits one simply needs to make sure the scene will fit in it
\begin{align}
n &= 0\;\text{m}   &  f &= 300\;\text{m} 
\end{align}
How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
\begin{align}
\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
\end{align}
and the distance from the image origin to its $y$-boundaries is given as
\begin{align}
\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
\end{align}
To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
\begin{align}
\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
\end{align}

@@

The ``camera image'' rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be. It does, however, show us what parts of the scene that are visible from the sonar. OpenGL can also produce a depth map that reveals the distance to each of image pixels. This information can be converted to a ranged sonar-like image by simply adding up all the intensity values that share the same depth for each range line. We perform this computation in OpenCL. It allows general purpose programming on GPUs and can interoperate with OpenGL quite nicely. This way we keep all the calculations on the GPU.

\begin{itemize}
\item Kernel tuning.
\item Sharing. PBO. Asynchronous.
\end{itemize}

@@

\subsection{Speed considerations}

Using OpenGL and OpenGL to generate sonar templates on the GPU allow hundreds of sonar templates to be formed per second. While this is sufficient for our needs. However, initializing the OpenGL and OpenCL context and loading 3D models from file into it is a slow process, and typically takes several seconds for every invocation. Therefore, to alleviate the simulation performance we either had to supply it with a list of parameters or ensure that it did not have to be shut down for every invocation.

The rendering loop that uses OpenGL and OpenCL to form

\begin{itemize}
\item Write to file
\end{itemize}
